---
title: "Getting started with coda.plot"
author: "Marc Comas-Cufí"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with coda.plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

`coda.plot` provides simple plots for exploring compositional data. In this type
of data, each row describes how a total is distributed among several parts. For
example, a mixture may consist of protein, fat, and carbohydrates.

The most important requirement is that all parts must be **strictly positive**.
Rows do not need to sum to 1 or 100: the functions work with relative
proportions.

```{r data}
library(coda.plot)

set.seed(2026)
X <- matrix(rexp(90, rate = 1), ncol = 3)
colnames(X) <- c("Protein", "Fat", "Carbohydrates")
group <- factor(rep(c("Control", "Treatment"), each = 15))

head(X)
```

## A quick ternary diagram

When there are exactly three parts, `ternary_plot()` is the most direct way to
display them. Each point corresponds to a row of `X`. A point closer to a vertex
has a larger relative proportion of the part shown at that vertex.

```{r ternary-plot}
ternary_plot(X, group = group)
```

With `center = TRUE`, the plot is centred on the mean composition. `scale = TRUE`
standardises variability, while `show_pc = TRUE` adds the first two principal
directions of variation.

```{r ternary-transformed}
ternary_plot(X, group = group, center = TRUE, scale = TRUE, show_pc = TRUE)
```

## Building a ternary diagram layer by layer

The modular interface is useful when you want to control each plot element.
First, `ternary_frame()` defines the transformation and labels. Next,
`ternary_base()` creates the triangle, and the `add_ternary_*()` functions add
layers to it.

```{r ternary-layers}
frame <- ternary_frame(X, labels = c("P", "F", "C"))

p <- ternary_base(frame, show_grid = FALSE)
p <- add_ternary_grid(p, ticks = c(0.25, 0.50, 0.75), colour = "grey80")
p <- add_ternary_points(p, X, group = group, size = 2)
p
```

`add_ternary_path()` joins ordered compositions. In this example, the path shows
a gradual transition from a protein-rich composition to a carbohydrate-rich
composition.

```{r ternary-path}
path <- rbind(
  c(8, 1, 1),
  c(6, 2, 2),
  c(4, 3, 3),
  c(2, 3, 5),
  c(1, 2, 7)
)

p <- ternary_base(ternary_frame(path, labels = colnames(X)))
p <- add_ternary_path(p, path, colour = "#0072B2", linewidth = 1)
add_ternary_points(p, path, colour = "#0072B2", size = 2)
```

`add_ternary_pc()` adds principal directions manually. It is the modular
equivalent of `show_pc = TRUE`.

```{r ternary-pc}
p <- ternary_base(ternary_frame(X, center = TRUE))
p <- add_ternary_points(p, X, group = group)
add_ternary_pc(p, X, colour = "black", linewidth = 0.7)
```

Finally, `ternary_coords()` returns the coordinates used for plotting. This is
useful for preparing annotations or custom `ggplot2` layers.

```{r ternary-coords}
coords <- ternary_coords(frame, X, group = group)
head(coords)
```

## Comparing groups with geometric means

`geometric_mean_barplot()` compares parts across groups. Bars represent
deviations from the overall mean; `include_boxplot = TRUE` also shows the
variability among observations.

```{r geometric-means}
geometric_mean_barplot(X, group, include_boxplot = TRUE)
```

With `clr_scale = TRUE`, calculations use clr coordinates, which are appropriate
for interpreting relative differences among parts.

```{r geometric-means-clr}
geometric_mean_barplot(X, group, clr_scale = TRUE)
```

## CLR biplot

For compositions with three or more parts, `clr_biplot()` summarises observations
and parts in two dimensions. Nearby points represent similar observations; label
directions indicate which parts explain the variation.

```{r biplot-data}
X6 <- matrix(rexp(180), ncol = 6)
colnames(X6) <- paste0("Part_", 1:6)
```

```{r clr-biplot}
clr_biplot(X6, group = group)
```

The `"covariance"` type emphasises the covariance structure among parts, so the
lengths and angles of the variable rays are the main elements to interpret. The
`"form"` type emphasises the geometry of the observations, so distances between
points are the main elements to interpret.

```{r clr-form}
clr_biplot(X6, group = group, biplot_type = "form")
```

If you need to reuse the coordinates, `return_data = TRUE` returns the observation
data, variable data, and plot.

```{r clr-return}
result <- clr_biplot(X6, group = group, return_data = TRUE)
names(result)
```

## Log-contrast biplot

`logcontrast_biplot()` displays observations according to two user-defined
contrasts. Each column of `lc` must sum to zero. Here, the first axis compares
parts 1 and 2, while the second compares parts 3 and 4.

```{r logcontrast}
lc <- cbind(
  `Part 1 / Part 2` = c(1, -1, 0, 0, 0, 0),
  `Part 3 / Part 4` = c(0, 0, 1, -1, 0, 0)
)

logcontrast_biplot(X6, lc, group = group)
```

## Balance dendrogram

`balance_dendrogram()` helps interpret a balance basis. The `B` matrix describes
which parts are compared at each split. One can be generated automatically with
`coda.base::pb_basis()`.

```{r balance-dendrogram}
B <- coda.base::pb_basis(X6, method = "exact")
balance_dendrogram(X6, B, group = group)
```

## Function summary

| Function | Main use |
|---|---|
| `ternary_plot()` | Quickly create a complete ternary diagram. |
| `ternary_frame()` | Define the ternary transformation and labels. |
| `ternary_base()` | Create the base triangle. |
| `add_ternary_grid()` | Add grid lines. |
| `add_ternary_points()` | Add observations. |
| `add_ternary_path()` | Add a path of ordered compositions. |
| `add_ternary_pc()` | Add principal component directions. |
| `ternary_coords()` | Obtain coordinates for custom layers. |
| `geometric_mean_barplot()` | Compare parts and groups with geometric means. |
| `clr_biplot()` | Explore observations and parts in clr coordinates. |
| `logcontrast_biplot()` | Display two user-defined log-contrasts. |
| `balance_dendrogram()` | Interpret a balance basis. |

To get started, `ternary_plot()` is usually enough for three-part compositions,
while `clr_biplot()` is a good choice for higher-dimensional compositions. The
remaining functions offer more control or answer more specific questions.
